How to Avoid OOTA Without Really Trying

### 2024-09-01 paper/ootavacuous.tex
paper/ootavacuous: Remove "Interthread Communications" appendix
Also the one reference to it.  We might want to add a citation defining
rf, rfi, and rfe, so added a comment to this effect.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Semantic dependencies form only one type of link in an OOTA cycle.
The other type extends from a given store to a load that returns the
value stored.
It is tempting to argue that the store must precede
the load in global time and combine this with the intuitive notion that
any real C++ implementation must consume global time when computing a
semantic dependency.
This combination suggests that OOTA cycles cannot occur.
The idea has been formalized by defining an OOTA cycle as a cycle
in sdep $\cup$ rf~\cite{PaulEMcKenney2014OOTA},
% This citation credits the formulation to Ali Sezgin, third author
% of N4323 ("Out-of-Thin-Air Execution is Vacuous").
where sdep is the set of semantic dependencies within
each thread and rf is the set of store-to-load ``reads-from'' links,
whether within a thread
(rfi) or between threads (rfe).\footnote{
	See Appendix~\ref{app:Interthread Communications} for definitions
	and properties of rf, rfe, and rfi.}

@@

\section{Interthread Communications}
\label{app:Interthread Communications}

This section more precisely defines the ``coe'', ``fre'', and ``rfe''
types of interthread communication (also collectively called ``links'')
and presents information about their temporal qualities.

First, naming.
The final ``e'' in all three acronyms stands for ``external'' (that
is, between threads) as opposed to ``internal'' communication (within
a thread: ``coi'', ``fri'', and ``rfi'') and also as opposed to all
communication, whether interthread or intrathread (``co'', ``fr'', and
``rf'').

Next, definitions:
\begin{description}
\item[coe]
	Coherence-order (external), which connects a store to any other store
	(in a different thread)
	to the same object that comes later in the object's
	modification order, i.e., that overwrites either the first store's
	value or some later value.
\item[fre]
	From-read (external), which connects a load to any store
	(in a different thread) to
	the same object that overwrites either the value that the load
	returned or some later value.
\item[rfe]
	Reads-from (external), which connects a store to any load
	(in a different thread) from
	the same object that returns the value stored.
	(Note: This means that the load retrieves the information
	written by the store, as opposed to retrieving the information
	from a different store that happened to write the same value.)
\end{description}

\begin{figure}[tb]
\begin{center}
\resizebox{4in}{!}{\includegraphics{ipc}}
\caption{ITC Diagram for coe, fre, and rfe}
\label{fig:ITC Diagram for coe, fre, and rfe}
\end{center}
\end{figure}

Thirdly, Figure~\ref{fig:ITC Diagram for coe, fre, and rfe} illustrates
the coe, fre, fri, and rfe links for four threads with time advancing
from the top to the bottom of the figure.  Start with coe: Even though
Thread~1's store is later in global time than that of Thread~0, the
Thread~1 store comes first in \co{x}'s modification order, then Thread~0's
store, then those of Threads~2 and~3.
There is thus a coe link from Thread~1's store to every other store,
from Thread~0's store to those of Threads~2 and~3, and from Thread~2's
store to that of Thread~3.
This situation can occur due to the initial placement and subsequent
movement of cache lines.

Thread~3's load reads the value written by Thread~0's store, so there is
an rfe link from that store to that load.  In addition there is an fre link from
the load to Thread~2's earlier store and an fri link to Thread~3's later store.
Note that the fre link goes backwards in time.

Finally, the reader might desire hard evidence that coe and fre
links really can go backwards in time.
We provide this evidence on x86 to demonstrate that these effects
are not confined to weakly ordered architectures.
The machine is a dual-socket system with
Intel(R) Xeon(R) Gold 6138 CPUs @ 2.00~GHz,
each socket with 20~cores and each core having a pair of hardware
threads, for a grand total of 80 hardware threads.
The code generating this data may be found in the \path{CodeSamples/cpu}
directory of the git archive of the book
``Is Parallel Programming Hard, And, If So, What Can You Do About
It?''~\cite{McKenney2018ParallelProgramming-2018-12-08a}.\footnote{
	\co{git://git.kernel.org/pub/scm/linux/kernel/git/paulmck/perfbook.git}}
See especially the \path{perftemporal.sh} script.

\begin{figure}[tb]
\begin{center}
\resizebox{4in}{!}{\includegraphics{coe-sh-out}}
\caption{On x86, coe Links Are Atemporal}
\label{fig:On x86; coe Links Are Atemporal}
\end{center}
\end{figure}

Figure~\ref{fig:On x86; coe Links Are Atemporal}
shows that coe links really are atemporal, that is, they can go backwards
in time, and quite frequently at that, even on x86.
During a run of the test, each of the threads stores a distinct value to
the same atomic integer variable (all at about the same time), and each
one records timestamps just before and just after its store.
The ``winning'' store is the one whose value is retained in the
variable at the end of the run, overwriting all the others.
The store-to-store latency is measured from the beginning of the
nonwinning store that started latest to the end of the winning store.
Every data point on the negative x-axis thus represents a group of
runs in which a nonwinning store started after the winning store
finished.
This demonstrates that the winning store quite commonly is not the
last one, even on x86.

\begin{figure}[tb]
\begin{center}
\resizebox{5in}{!}{\includegraphics{coe-out-1}}
\caption{On x86, coe Links Form a Coherent Partial Order}
\label{fig:On x86, coe Links Form a Coherent Partial Order}
\end{center}
\end{figure}

Figure~\ref{fig:On x86, coe Links Form a Coherent Partial Order} shows
that the sequence of values read from the atomic variable by each
thread is consistent with a number of global orders, as
required.\footnote{
	Note that this data was not measured on the 80-thread system but
	rather on a 16-thread x86 laptop, to avoid an unreadable diagram
	containing 79 bubbles.}
Working from the far left, one thread saw the values 0, 11, and 9,
a second thread saw 3, 11, and 9, and so on.
Although different threads saw different sequences of values, there is no
disagreement on the order of the values that they could see.
For example, all threads agree that the value 9 came last.
This is an example of single-variable sequential consistency, and it
meets the coherence requirements of the C++ standard.

\begin{figure}[tb]
\begin{center}
\resizebox{3in}{!}{\includegraphics{fre-sh-out}}
\caption{On x86, fre Links Are Atemporal}
\label{fig:On x86, fre Links Are Atemporal}
\end{center}
\end{figure}

Figure~\ref{fig:On x86, fre Links Are Atemporal}
plots a histogram of the elapsed time from the beginning of a load that
returned an old value to the end of the store that provided the new value.
Most of the data falls into negative time.
In fact, in the most commonly occurring case the last load of an old value
executes about 60 timestamp periods (about 30~nanoseconds) \emph{after}
the store which overwrote that old value.
This shows that fre links can and do go backwards in global time,
corresponding to the fre link between Thread~3 and Thread~2 in
Figure~\ref{fig:ITC Diagram for coe, fre, and rfe}.

\begin{figure}[tb]
\begin{center}
\resizebox{3in}{!}{\includegraphics{rfe-sh-out}}
\caption{On x86, rfe Links Are Temporal}
\label{fig:On x86, rfe Links Are Temporal}
\end{center}
\end{figure}

Figure~\ref{fig:On x86, rfe Links Are Temporal}
plots a histogram of the elapsed time from the store of a new value
to the first load of that new value.
Here all of the data falls into positive time, indicating
that rfe links always go forward in global time, as those familiar with
computer hardware, limits on speculative execution, and the laws of
physics would expect.\footnote{
	Our apologies to those who might feel that this is belaboring
	the obvious.
	And please understand that these points have proven helpful to
	some of those whose work has never strayed too close to
	hardware.}

\begin{figure}[tb]
\begin{center}
\resizebox{4in}{!}{\includegraphics{coe}}
\resizebox{4in}{!}{\includegraphics{fre}}
\resizebox{4in}{!}{\includegraphics{rfe}}
\caption{Propagation Delay and Temporal Properties of coe, fre, and rfe}
\label{fig:Propagation Delay and Temporal Properties of coe, fre, and rfe}
\end{center}
\end{figure}

Figure~\ref{fig:Propagation Delay and Temporal Properties of coe, fre, and rfe}
shows how propagation delay explains the temporal properties of coe,
fre, and rfe.
The upper portion of the figure shows an atemporal coe, in which
CPU~0 stores 1 to \co{x} after CPU~3 stores 2 but
the modification order of \co{x} is decided after the fact.
In this case, the value of 2 from CPU~3's store overwrites the value
of 1 from CPU~0's store, despite the fact that CPU~0's store
happened later in global time.
One way this could happen is if the cache line containing
\co{x} arrived at CPU~0 before arriving at CPU~3.
A more fanciful explanation is that both values arrived at an
interconnect at the same time, and the interconnect made an
arbitrary choice between the two.
Either way, the end result is that an earlier store can overwrite
a later store.

The middle portion of the figure shows an atemporal fre, in which
CPU~3's load obtains the old value of zero from \co{x} despite
having executed long after CPU~0's store, courtesy of the fact
that the new value of \co{x} had not yet propagated to CPU~3.

The lower portion of the figure shows a temporal rfe.
Because a load on CPU~3 cannot obtain the value stored by CPU~0 until
after that value has propagated to CPU~3, the fact that CPU~3 obtains
the new value implies that the load was executed after the store.

And this is exactly why the C++ memory model guarantees ordering from
rfe links but not from coe and fre links for relaxed accesses (in the
absence of other ordering from stronger atomic memory accesses or
\co{atomic_thread_fence(memory_order_seq_cst)}).

>>>>

Semantic dependencies form only one type of link in an OOTA cycle.
The other type extends from a given store to a load that returns the
value stored.
It is tempting to argue that the store must precede
the load in global time and combine this with the intuitive notion that
any real C++ implementation must consume global time when computing a
semantic dependency.
This combination suggests that OOTA cycles cannot occur.
The idea has been formalized by defining an OOTA cycle as a cycle
in sdep $\cup$ rf~\cite{PaulEMcKenney2014OOTA},
% This citation credits the formulation to Ali Sezgin, third author
% of N4323 ("Out-of-Thin-Air Execution is Vacuous").
where sdep is the set of semantic dependencies within
each thread and rf is the set of store-to-load ``reads-from'' links,
whether within a thread (rfi) or between threads (rfe).
% @@@ Consider adding citation defining rf, rfi, and rfe.

@@
